---
title: "Predict Price of Diamonds"
author: "Ajay Shankar A"
date: "2023-11-14"
format: html
code-fold: false
categories: [Analysis, Code, EDA, Modeling]
---

## Predicting Diamonds Price

### Introduction

Building a model to **predict the price of the diamonds** using `tidymodels`.

  Diamonds data set is readily available to use through the `ggplot2` library in the `tidyverse` and we will be using this data set predict the prices of the other diamonds.

In the data set various parameters of diamonds are given and each of these parameters may or may not effect the **price** of the diamonds.


```{r setup}
#| echo: false

library(knitr)

knitr::opts_chunk$set(warning = FALSE, echo = TRUE, dpi = 150,
                      fig.height = 5, fig.width = 8)

```

```{r loading libraries and data}
#| message: false
#| warning: false

library(tidyverse)

data("diamonds")
diamonds

```
Data has over 50,000 observations which is good for modeling.

### Exploring the data

The diamonds data set is available to explore in `ggplot2` library as mentioned above.

Let's check for NA's before exploring the data

```{r na_check}

diamonds %>% map( ~sum(is.na(.))) %>% unlist()

```



```{r price_histogram}
#| label: fig-price_hist

diamonds %>% ggplot(aes(carat)) + geom_freqpoly(binwidth = 0.05)

```
From the @fig-price_hist we can observe that
- Most of the diamonds are between 0.2 to 1.5 carats.
- There are peaks which means higher number of diamonds at whole and common fractions.

My general knowledge is that the weight i.e, carat of the diamond influences the price most. Let's visualize that.


```{r price_vs_carat}

diamonds %>% ggplot(aes(carat, price)) + geom_hex(bins = 50)

```

The price tends to follow exponential curve the `log2()` curve, we can confirm this by another graph.

```{r log_price_vs_carat}
#| label: fig-log_prc_vs_carat

diamonds %>% filter(carat < 2.5) %>% 
  mutate(log_price = log2(price),
                    log_carat = log2(carat)) %>% 
  ggplot(aes(log_carat, log_price)) + geom_hex(bins = 50)
```

The above @fig-log_prc_vs_carat shows that once we apply `log2()` to both price and carat the relationship mostly looks to be linear.

```{r}
diamonds %>% filter(carat <= 2.5) %>% ggplot(aes(carat, price)) +
  geom_col()
```

We can see that `price` jumps when the weight is exactly or greater than to the whole and common fractions such as 0.5, 1.0, 1.5 and 2.


```{r exploring the data}
library(patchwork)

plot_parameter <- function(param){
  ggplot(diamonds, aes({{param}}, price)) + geom_boxplot()
}

(plot_parameter(cut) + plot_parameter(color)) /
  (plot_parameter(clarity))

```





